---
title: "Data Science Coursework 2"
author: "06026373"
date: "2025-05-10"
output: pdf_document
---


<style type="text/css">
h1{
  font-size: 24pt;
}
h2{
  font-size: 18pt;
}
body{
  font-size: 12pt;
}
</style>

```{r setup, include = FALSE, tidy=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
include_solutions <- TRUE
```
```{r setup2, include=FALSE, tidy=TRUE}
require(rmarkdown)
require(knitr)
require(kableExtra)
require(data.table)
require(dplyr)
require(ggplot2)
require(cmdstanr)
require(bayesplot)
```

# Academic integrity statement

"I, 06026373, certify that this assessed coursework is my own work,
unless otherwise acknowledged, and includes no plagiarism. I 
have not discussed my coursework with anyone else except when seeking
clarification with the module lecturer via email or on MS Teams. I have not
shared any code underlying my coursework with anyone else prior to submission."

# Part 1 - Introduction

In the last decades, a statistic called "[expected Goal](https://en.wikipedia.org/wiki/Expected_goals)" (xG), has been developed [...]. It represents the likelihood of a shot being scored and is based on several factor like the distance to the goal, the body part used, the distance to the closest defender, \dots It has been constantly improved by new studies during the last decades. However, this likelihood is not just due to luck: the difference in individual competence of each player also plays a significant role. Some players consistently outperform their expected goals, suggesting superior finishing skills, decision-making under pressure, \dots.

This project aims to see if a player scoring more or less than there expected goal can be due to a personal competence of simply luck.

# Part 2 - Data

```{r data, echo = FALSE, tidy = TRUE}
################
# Part 2 - Data#
################

data.dir <- 'data/'
out.dir <- 'output/'
```

We load the data from the website [FBREF](https://fbref.com), which contains a lot of data about each matchs in several leagues. We will use the data from the Premier League, saison 23/24 and 22/23. This data contains a lot of information, however we will only be concerned about: players id, match id, team id, opposite team id, the number of goal of the player in the match, the number of shots, the expected Goal for the player for the match.

```{r data2, echo = FALSE, tidy = TRUE}
# Import the data
file_2324 <- file.path(data.dir, 'Premier_League2324.csv')
file_2223 <- file.path(data.dir, 'Premier_League2223.csv')
df_2324 <- as.data.table(read.csv(file_2324))
df_2223 <- as.data.table(read.csv(file_2223))

# Take only the relevant column:
# player.tag, team.tag, id_team_A, id_team_B, match.tag, Goals, Shots.Total, xG..Expected.Goals, # Minutes:
df_2324 <- df_2324[, .('Player' = player.tag,
                       'Team' = team.tag,
                       'Team_A' = id_team_A,
                       'Team_B' = id_team_B,
                       'Match' = match.tag,
                       'Goals' = Goals,
                       'Shots' = Shots.Total,
                       'xG' = xG..Expected.Goals,
                       'Minutes' = Minutes
                        )]
df_2223 <- df_2223[, .('Player' = player.tag,
                       'Team' = team.tag,
                       'Team_A' = id_team_A,
                       'Team_B' = id_team_B,
                       'Match' = match.tag,
                       'Goals' = Goals,
                       'Shots' = Shots.Total,
                       'xG' = xG..Expected.Goals,
                       'Minutes' = Minutes
                        )]

# Combine both datasets
df <- bind_rows(df_2223, df_2324)

# Create an Opposite_Team column
df <- df %>%
  mutate(
    Opposite_Team = ifelse(Team == Team_A, Team_B, Team_A)
  )
```

We will only keep players that have scored at least 5 goals, or have at least 5 as xG. With this, we are guarantied to have data on players focusing on goals, while still having enough players to analyse:
```{r data3, echo = FALSE, tidy = TRUE}
# Create a dataset with only players scoring at least 5 goals or with xG >=5
df_combined <- df %>%
  group_by(Player) %>%
  summarise(
    total_minutes = sum(Minutes, na.rm = TRUE),
    total_goals = sum(Goals, na.rm = TRUE),
    total_xG = sum(xG, na.rm = TRUE)
  ) %>%
  filter(total_goals >= 5 |
           total_xG >= 5)

# Filter the original datasets
df <- df %>%
  filter(Player %in% df_combined$Player)

cat('There are ')
cat(nrow(df))
cat(' entries, composed of ')
cat(length(unique(df$Player)))
cat(' unique players.')
```
# Part 3 - Exploratory data analysis

Before any analysis, we will just check the difference between the number of goal of each players and there expected goal. Normally this value should be close to 0 if the expected goal value is well constructed. 

```{r exploratory, echo = FALSE, tidy = TRUE}
######################################
# Part 3 - Exploratory data analysis #
######################################

df_combined$diff <- df_combined$total_goals - df_combined$total_xG

cat('The mean of the difference between the goal scored and xG is: ')
cat(mean(df_combined$diff))
cat('\nThe variance of the difference between the goal scored and xG is: ')
cat(var(df_combined$diff))
```
```{r exploratory2, echo = FALSE, tidy = TRUE}
# Create a graphique to plot the difference
ggplot(df_combined, aes(x = diff)) +
  geom_histogram(aes(y = ..density..), binwidth = 1,
                 fill = "steelblue", color = "black") +
  stat_function(fun = dnorm, args = list(mean = 0, sd = sqrt(10.5)), color = "red", size = 1) +
  labs(
    title = "Distribution de la différence stat_1 - stat_2",
    x = "Difference",
    y = "Frequency"
  ) +
  theme_minimal()
```
We can see that the mean is closed to 0, this suggest that the expected Goal values is a good tool for this analysis. The data is also closed to a normal law, which confirm the robustnest of this statistics. However the variance is only 10.5, which is not a lot as players in the data base played around 50 matchs, so we have only a \pm 0.06 difference. But this is still a lot as there is 38 matchs in a saison, this may result in a difference of 3 goals for the whole season between the players with the best difference, and the worse one. 

So it is interesting to see if the difference between the number of goals and the expected goal is due to luck or personal competence.

# Part 4 - Multi-models analysis

## Theoretical model

To quantify the difference between luck and competence, we will use the following Bayesian hierarchical model:
$$Y_{i,j} \sim \text{Bernouilli}(p_{i,j})$$
$$\text{logit}(p_{i,j})=\text{logit}(\text{xG}_{i,j})+\theta_i$$
$$\theta_i \sim \mathcal{N}(0,\sigma^2)$$
$$\sigma \sim \text{Exp}(1)$$
With $Y_{i,j} \in {0,1}$ the categorical variable that take the value 1 if the $j$ shot by the $i$ players was a goal, $\theta_i$ the competence of the player $i$, $\text{xG}_{i,j}$ the expected goal of the $j$ shot by the $i$ players.

However we do not have the expected goal statistic for each shot, we only have the expected goal for the whole match. So we will use the following model instead:
$$Z_{i,k} \sim \text{Bin}(n_{i,k}, q_{i,k})$$

$$\text{logit}(q_{i,k})=\text{logit} \left (\frac{\text{xG}_{i,k}}{n_{i,k}} \right )+\theta_i$$
$$\theta_i \sim \mathcal{N}(0,\sigma^2)$$
$$\sigma \sim \text{Exp}(1)$$
With $Z_{i,k}$ the number of goal scored by the player $i$ on match $k$, $n_{i,k}$ the number of shot taken by the player $i$ on match $k$, $\text{xG}_{i,k}$ the expected goal for the whole match $k$.

```{r multi-model, echo = FALSE, tidy = TRUE}
##################################
# Part 4 - Multi-models analysis #
##################################

# Stan code
competence_txt <- "
data {
  int<lower=1> N;  // number of entries
  int<lower=1> P;  // number of players
  array[N] int<lower=0> Z;  // number of goals scored in the match
  array[N] int<lower=0> S;  // number of shots in the match
  vector[N] mean_xG;  //mean of the expected goals of the shots in the match = xG / T
  array[N] int<lower=1> player_id;
}
parameters {
  vector[P] theta;
  real<lower=0> sigma;
}
transformed parameters {
  vector[N] q;
  
  q = inv_logit(logit(mean_xG) + theta[player_id]);
}
model {
  // Likelihood
  Z ~ binomial(S, q);
  // Priors
  theta ~ normal(0, sigma);
  sigma ~ exponential(1);
}
"
```
```{r multi-model2, echo = FALSE, tidy = TRUE}
# data preparation
df$xG_mean <- df$xG / df$Shots
df <- df %>%
  filter(!is.nan(xG_mean)) # Remove lines with no shots
df <- df %>%
  filter(xG_mean <= 1) # Remove lines where xG > shots
df <- df %>%
  mutate(xG_mean = ifelse(xG_mean == 1, 0.999,
                   ifelse(xG_mean == 0, 0.001, xG_mean))) # to avoid problem with inv_logit
df <- df %>%
  mutate(Shots = pmax(Shots, Goals))
df$player_id = as.integer(factor(df$Player)) # Create id for the players

```
```{r multi-model3, echo = FALSE, tidy = TRUE}
# compile the model
competence_filename <- cmdstanr::write_stan_file(
  gsub('\t',' ', competence_txt ),
  dir = out.dir,
  basename = NULL,
  force_overwrite = FALSE,
  hash_salt = ""
)
competence_compiled <- cmdstanr::cmdstan_model(competence_filename)

# data
stan_data <- list(
  N = nrow(df),
  P = length(unique(df$player_id)),
  Z = df$Goals,
  S = df$Shots,
  mean_xG = df$xG_mean,
  player_id = df$player_id
)

# sample
competence_fit <- competence_compiled$sample(
  data = stan_data,
  seed = 6373,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 500,
  iter_sampling = 4500,
  refresh = 500
)

```
## Checking convergence

Before analysing the results, we check if the model converges well.

```{r check, echo = FALSE, tidy = TRUE}
# Check the model
check <- competence_fit$summary(
  variables = c('theta','sigma'),
  posterior::default_summary_measures(),
  posterior::default_convergence_measures()
)

cat('The maximum of rhat is: ')
cat(max(check$rhat))
cat('\nThe minimum of ess_bulk is: ')
cat(min(check$ess_bulk))
cat('\nThe minimum of ess_tail is: ')
cat(min(check$ess_tail))

```
We have rhat < 1.1 and ess > 100. So the model seems to converge. Let's check y ploting the "worst" variable:

```{r check2, echo = FALSE, tidy = TRUE, fig.align='center'}
# parameter with lowest ess_bulk
competence_worst_var <- check$variable[which.min(check$ess_bulk)]

# extract samples
competence_po <- competence_fit$draws(
  variables = c("lp__",competence_worst_var),
  inc_warmup = FALSE,
  format = "draws_array"
  )

# make trace plot
p <- bayesplot:::mcmc_trace(competence_po,  
                            pars = c("lp__",competence_worst_var),
                            facet_args = list(nrow = 2)
                            )
p <- p + theme_bw()
ggsave(file = file.path(out.dir,'competence_worst_trace.png'), 
       plot = p, 
       h = 10, 
       w = 8
       ) # change

p

```
Even if we see some spikes, the variable converges. So our model converges.

## Results

We can now

```{r analysis, echo = FALSE, tidy = TRUE}
competence_po <- competence_fit$draws(
  variables = c("theta"),
  inc_warmup = FALSE,
  format = "draws_df"
  )

competence_po <- as.data.table(competence_po)
competence_po <- data.table::melt(competence_po, id.vars = c('.chain','.iteration','.draw') )

competence_po_summary <- 
  competence_po[,
            list( summary_value = quantile(value, prob = c(0.025, 0.25, 0.5, 0.75, 0.975)),
                  summary_name = c('q_lower','iqr_lower','median','iqr_upper','q_upper') 
                  ),
            by = 'variable'
            ]
```


# Code appendix  

```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```
